Off the beaten path: \\ Gaia reveals GD-1 stars outside of the main stream

### 2018-05-01 paper/gd1.tex
width

Clear density variations, also apparent in the 2D positions of the sources,
manifest as sharp features in the surface density estimates along the stream.
These do not map to features in dust extinction nor are they caused by completeness variations
in the \gaia\ scanning pattern, as 99\% of our sample has at least 8 \gaia\ visibility periods,
promising a robust estimate of the proper motions.
% First, the visible portions of the GD-1 stream are located at high Galactic
% latitudes ($b > 20^\circ$).
%, and we therefore do not expect significant dust
% extinction or variations in extinction within the stream region.
% The median $V$-band extinction over the stream track is
% $\textrm{med}\left(A_V\right) = 0.04~\textrm{mag}$ with a dispersion
% $\sigma_{A_V} \approx 0.03~\textrm{mag}$, as computed from the SFD extinction
% map (\citealt{Schlegel:1998}).
% As the stream approaches the Galactic disk ($\phi_1 < -60^\circ$),
% dust extinction becomes appreciable, and the stream stars are harder to
% select apart from the background.
% Second, 99\% of the stars in our sample have at least 8 \gaia\ visibility
% periods (median and root-variance of 14 and 3, respectively), promising a robust
% estimate of the proper motions.
% The variance in the number density of faint stars originating from the \gaia\
% scanning pattern is only $\approx 1~\textrm{deg}^{-2}$, as estimated using stars
% with proper motions $-9 < \mu_{\phi_1, \odot} < -4.5~\masyr$ and $1 <
% \mu_{\phi_2, \odot} < 4~\masyr$, far smaller than the density variations in the
% stream.
% The relative ratios between different over- and under-densities might be affected by the \gaia\ completeness at the faint end, but they
The observed density variations are therefore likely real morphological changes
along the GD-1 stream.
% as have been suggested before photometric selection alone
% (\citealt{Carlberg:2013, DeBoer:2018}).
The deep under-densities we see from the kinematically-cleaned sample correspond
to previously reported gaps in the stream (\citealt{Carlberg:2013, DeBoer:2018}).
There is also a well-defined, sharp overdensity in the surface density profile
close to $\phi_1 \approx -13^\circ$ with roughly symmetric under-densities on
either side:
this is suggestive of a progenitor system in the final stages of dissolution
(\citealt{Balbinot:2018}).
This region is also the thinnest part of the stream, as would be expected for
recently stripped material.
% that has not yet phase-mixed in multiple dimensions.

@@

To initialize the model stream generation, we assume that the present day
position of the progenitor system is at $\phi_1 = -13.5^\circ$, the peak of the
observed surface density profile.
We numerically integrate the backwards from this location for $4~\textrm{Gyr}$,
then forward-generate the model stream, and release stream particles at each
timestep.
We assume an initial progenitor mass of $M=10^5~\msun$, and linearly decrease
the mass of the progenitor until its full disruption at the present.
% day, where we assume $M = 0~\msun$
% (i.e. fully disrupted).
Stream properties computed from this resulting model stream are plotted in
\figurename~\ref{fig:track-and-model} as dashed (orange) lines.
The model stream qualitatively matches the stream track in all phase-space
dimensions, but has a much smoother surface density profile along the extent of
the stream.
We therefore don't expect that the observed surface density variations can be fully explained
by, e.g., epicyclic overdensities (\citealt{Kupper:2012}).

>>>>

Clear density variations, also apparent in the 2D positions of the sources,
manifest as sharp features in the surface density estimates along the stream.
These do not map to features in dust extinction nor are they caused by completeness variations
in the \gaia\ scanning pattern, as 99\% of our sample has at least 8 \gaia\ visibility periods,
promising a robust estimate of the proper motions.
% First, the visible portions of the GD-1 stream are located at high Galactic
% latitudes ($b > 20^\circ$).
%, and we therefore do not expect significant dust
% extinction or variations in extinction within the stream region.
% The median $V$-band extinction over the stream track is
% $\textrm{med}\left(A_V\right) = 0.04~\textrm{mag}$ with a dispersion
% $\sigma_{A_V} \approx 0.03~\textrm{mag}$, as computed from the SFD extinction
% map (\citealt{Schlegel:1998}).
% As the stream approaches the Galactic disk ($\phi_1 < -60^\circ$),
% dust extinction becomes appreciable, and the stream stars are harder to
% select apart from the background.
% Second, 99\% of the stars in our sample have at least 8 \gaia\ visibility
% periods (median and root-variance of 14 and 3, respectively), promising a robust
% estimate of the proper motions.
% The variance in the number density of faint stars originating from the \gaia\
% scanning pattern is only $\approx 1~\textrm{deg}^{-2}$, as estimated using stars
% with proper motions $-9 < \mu_{\phi_1, \odot} < -4.5~\masyr$ and $1 <
% \mu_{\phi_2, \odot} < 4~\masyr$, far smaller than the density variations in the
% stream.
% The relative ratios between different over- and under-densities might be affected by the \gaia\ completeness at the faint end, but they
The observed density variations are therefore likely real morphological changes
along the GD-1 stream.
% as have been suggested before photometric selection alone
% (\citealt{Carlberg:2013, DeBoer:2018}).
The deep under-densities we see from the kinematically-cleaned sample correspond
to previously reported gaps in the stream (\citealt{Carlberg:2013, DeBoer:2018}).
There is also a well-defined, sharp overdensity in the surface density profile
close to $\phi_1 \approx -13^\circ$ with roughly symmetric under-densities on
either side:
this is suggestive of a progenitor system in the final stages of dissolution
(\citealt{Balbinot:2018}).
This region is also the thinnest part of the stream, as would be expected for
recently stripped material, but we defer a robust analysis of variations in the
stream width to future work.
% that has not yet phase-mixed in multiple dimensions.

@@

To initialize the model stream generation, we assume that the present day
position of the progenitor system is at $\phi_1 = -13.5^\circ$, the peak of the
observed surface density profile.
We numerically integrate the backwards from this location for $4~\textrm{Gyr}$,
then forward-generate the model stream, and release stream particles at each
timestep.
We assume an initial progenitor mass of $M=10^5~\msun$, and linearly decrease
the mass of the progenitor until its full disruption at the present.
% day, where we assume $M = 0~\msun$
% (i.e. fully disrupted).
Stream properties computed from this resulting model stream are plotted in
\figurename~\ref{fig:track-and-model} as dashed (orange) lines.
The model stream qualitatively matches the stream track in all phase-space
dimensions, but has a much smoother surface density profile along the extent of
the stream.
The broadening in the observed stream at negative longitudes is also seen in the model stream and is expected for older parts of the stream that have phase-mixed for longer times.
We therefore don't expect that the observed surface density variations can be fully explained
by, e.g., epicyclic overdensities (\citealt{Kupper:2012}).
